---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# betwixt

<!-- badges: start -->

[![lifecycle](https://lifecycle.r-lib.org/articles/figures/lifecycle-experimental.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental) [![Project Status: WIP](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip) [![devel-version](https://img.shields.io/badge/devel%20version-0.0.4-blue.svg)](https://github.com/dataobservatory-eu/betwixt) [![dataobservatory](https://img.shields.io/badge/ecosystem-dataobservatory.eu-3EA135.svg)](https://dataobservatory.eu/)

<!-- badges: end -->

Betwixt is a lightweight framework for constructing, representing, and reviewing candidate semantic claims in tabular data. It provides a pragmatic review layer in which candidate knowledge can be organised into bounded review tasks and presented for human judgement.

Betwixt keeps candidate semantic structure separate from its presentation for review. Candidate data can be projected into human-reviewable forms, while review decisions, states, and provenance remain explicit and reproducible.

## Installation

You can install the development version of betwixt from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("dataobservatory-eu/betwixt")
```

## Example

The `delini` dataset provides a small cultural heritage example containing
candidate semantic assertions prepared for human review:

```{r example}
library(betwixt)

data("delini")
delini
```


```{r render}
betwixt_render(
  delini,
  cols = c(
    col_1 = "Subject",
    col_2 = "instance of",
    col_3 = "heritage of",
    context_1 = "held by"
  ),
  subheadings = c(
    col_2 = "wdt:P31",
    col_3 = "controlled range"
  ),
  title = "Delini semantic review",
  description = "Review the proposed semantic assertions.",
  project_id = "delini",
  filename_stem = "delini-wide",
  sequence = 0L,
  path = tempdir()
)
```

The resulting standalone HTML review presents the evidence and descriptive context alongside the candidate semantic assertions and records the resulting review state and provenance.

## Vignettes

Betwixt implements the following workflow:

```
candidate knowledge
        ↓
bounded review task
        ↓
review projection
        ↓
human review
        ↓
review state
```
It is organised around complementary vignettes.

### 1. Betwixt Implementation

Introduces scoped claims, semantic stabilisation, contextual inheritance, and the conceptual foundations of Betwixt.

### 2. Review Layouts and Semantic Projections

Introduces the Delini example and shows how candidate semantic material can be represented as wide, dual-wide, long, and dual-long projections. The current browser-based review implementation uses the wide projection.

### 3. R Reference Implementation

Describes the R implementation and the construction and processing of candidate semantic data.

### 4. Python Minimal Implementation

Planned vignette demonstrating how the underlying approach can be implemented outside R using ordinary tabular and web technologies.
