---
title: "Review Layouts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Review Layouts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(betwixt)
```

# Review layouts and semantic projections

Betwixt supports the review of candidate semantic assertions before they are
accepted into a stabilised semantic state. The same candidate knowledge can be
presented in different tabular projections depending on what the reviewer needs
to inspect.

This vignette introduces these projections through the `delini` example
dataset. It distinguishes four related forms:

- wide;
- dual-wide;
- long; and
- dual-long.

These are not four different semantic models. They are different projections
of the same underlying candidate assertions and their supporting information.

The wide projection is currently the primary browser-based review interface in
Betwixt. The long projections are introduced here because they make the atomic
claim structure explicit and clarify how the review representation relates to
semantic graph statements.

## The Delini example

The `delini` dataset provides a small heterogeneous cultural heritage example
associated with the Delini farmstead at the Ethnographic Open-Air Museum of
Latvia. It contains **three artefacts and two archival records**: the farmhouse,
a tablet-woven sash, a bed, a record containing photographs of the farmhouse,
and a floor plan prepared for its reconstruction.

The example combines four kinds of information required for semantic review:
evidence, human-readable description, candidate semantic assertions, and
display-only context.

```{r delini-dataset}
data("delini")

delini
```

The columns belong to four functional groups:

| Role | Columns | Purpose |
|---|---|---|
| Evidence | `evidence_url`, `evidence_text` | Identifies the evidence presented to the reviewer |
| Description | `label`, `description` | Identifies and describes the resource in human-readable form |
| Candidate assertions | `col_1`, `col_2`, `col_3` and their range and definition columns | Contains the semantic assertions proposed for review |
| Context | `context_1` | Provides information outside the current review boundary |

In this review task, `col_1` identifies the **subject**, `col_2` proposes its
type through **`instance of`**, and `col_3` proposes its **`heritage of`**
association. The generic column names keep the candidate data structure
independent of a particular ontology or presentation vocabulary.

Candidate columns may be accompanied by a `_range` and a `_definition`.
Ranges provide controlled or suggested values, while definitions provide
resolvable semantic references where available.

The `delini` dataset is already in the **wide projection**: each row brings
together evidence, description, a subject, several candidate assertions, and
context. The other projections rearrange or extend this same candidate
semantic material while preserving these distinctions.

## Projections are views of candidate knowledge

A semantic review task rarely needs to expose every aspect of a graph in its
native graph representation. Betwixt therefore uses tabular projections to
present a bounded part of candidate knowledge for a particular review task.

A projection changes how candidate knowledge is arranged for inspection. It
does not, by itself, change what the assertions mean.

The Delini example can be represented in four closely related forms.

## Wide projection

The wide projection places a subject and several candidate assertions about it
on the same row:

```text
Evidence | Subject | instance of | heritage of | Context
```

Predicates are represented by columns and their candidate values by cells. This
keeps related assertions visible together and is particularly suitable for
curators and domain experts who review a described resource rather than a
sequence of atomic graph statements.

The wide projection therefore hides some graph mechanics without discarding
semantic structure. It presents a bounded part of that structure in a
task-oriented form. The `delini` dataset is represented in this wide form.


## Dual-wide projection

The dual-wide projection adds a reviewable relation between the evidence and
the subject:

```text
Evidence | Evidence relation | Subject | instance of | heritage of | Context
```

This is useful when the presence of evidence alone does not establish how that
evidence relates to the subject. An image may, for example, **depict** an
artefact, while an archival resource may **document** it.

The evidence relation is a candidate assertion, not review provenance.
Provenance instead records the review activity itself, including the reviewer,
decisions, sequence, and timestamps.

The evidence relation is optional: ordinary wide review can present supporting
evidence without making its relation to the subject part of the review task.

The dual-wide projection can therefore be derived directly from `delini` by
adding the evidence relation and its controlled range:

```{r delini-dual-wide}
delini_dual_wide <- delini |>
  dplyr::mutate(
    evidence_relation = c(
      "depicts", "depicts", "depicts", "documents", "documents"
    ),
    evidence_relation_range = candidate_range(
      "depicts", "documents", "Other…"
    ),
    .after = evidence_text
  )

delini_dual_wide
```

The first three evidential resources depict the three artefacts, while the
last two document the two archival records. All other candidate assertions
remain unchanged.

## Long projection

The long projection makes the atomic semantic claim structure explicit.
Instead of representing predicates as columns, each row contains one
subject–predicate–value assertion:

```text
Evidence | Subject | Predicate | Value
```

The wide assertions concerning one subject therefore become multiple rows. In
the Delini example, each resource has candidate `instance of` and `heritage of`
assertions, so the five wide rows become ten atomic candidate claims.

The long projection can be derived directly from `delini`:

```{r delini-long}
delini_long <- delini |>
  tidyr::pivot_longer(
    cols = c(col_2, col_3),
    names_to = "candidate_column",
    values_to = "value"
  ) |>
  dplyr::mutate(
    predicate = dplyr::recode(
      candidate_column,
      col_2 = "instance of",
      col_3 = "heritage of"
    ),
    subject = col_1
  ) |>
  dplyr::select(
    row_number,
    evidence_url, evidence_text,
    label, description,
    subject, predicate, value,
    context_1
  )

delini_long
```

Each original row now produces two rows with the same subject: one for
`instance of` and one for `heritage of`. The semantic content is unchanged;
only its tabular projection has changed.

This representation makes the boundary of an individual assertion explicit
and is closer to the subject–predicate–value structure used in graph
serialization. It is therefore useful for semantic modelling, review-task
design, data stewardship, and inspection of atomic candidate claims.

For domain-expert review, however, repeatedly presenting the same resource
across several atomic rows may be less natural than seeing its related
properties together. Wide and long are therefore not merely cosmetic table
layouts: they provide different views of the same candidate semantic material
for different review tasks.


## Dual-long projection

The dual-long projection combines atomic semantic claims with an explicit,
reviewable relation between the evidence and the subject:

```text
Evidence | Evidence relation | Subject | Predicate | Value
```

It can be derived by pivoting the dual-wide projection in the same way as the
ordinary wide projection:

```{r delini-dual-long}
delini_dual_long <- delini_dual_wide |>
  tidyr::pivot_longer(
    cols = c(col_2, col_3),
    names_to = "candidate_column",
    values_to = "value"
  ) |>
  dplyr::mutate(
    predicate = dplyr::recode(
      candidate_column,
      col_2 = "instance of",
      col_3 = "heritage of"
    ),
    subject = col_1
  ) |>
  dplyr::select(
    row_number,
    evidence_url,
    evidence_text,
    evidence_relation,
    evidence_relation_range,
    label,
    description,
    subject,
    predicate,
    value,
    context_1
  )

delini_dual_long
```

Each row now exposes two distinct semantic relationships: the relation between
the evidence and the subject, and one atomic assertion about that subject. As
in the dual-wide projection, the evidence relation is a candidate assertion
and remains distinct from review provenance.

The four projections therefore preserve the same basic semantic material while
varying two dimensions: whether assertions are arranged in wide or atomic-long
form, and whether the evidence relation is implicit or explicitly reviewable.

## Relationship between the four projections

The four projections vary along two independent dimensions: **wide versus
long** determines how semantic assertions are arranged, while **ordinary
versus dual** determines whether the relation between evidence and subject is
itself reviewable.

| Projection | Assertion structure | Evidence relation |
|---|---|---|
| Wide | Multiple candidate predicates per row | Not reviewable |
| Dual-wide | Multiple candidate predicates per row | Reviewable |
| Long | One atomic candidate claim per row | Not reviewable |
| Dual-long | One atomic candidate claim per row | Reviewable |

Moving from wide to long rearranges the same candidate assertions into atomic
subject–predicate–value rows. Moving from ordinary to dual adds an explicit
candidate assertion relating the evidence to the subject.

These projections therefore separate the **semantic content of the candidate
knowledge** from its **representation for a particular review task**. The same
candidate material can be projected differently without making one tabular
representation canonical.

## From review design to review execution

Long and wide projections support different stages of semantic review. The long
form exposes atomic subjects, predicates, values, ranges, and definitions,
making it useful for semantic modelling and review-task design. Once the
relevant claims and permitted choices have been bounded, they can be presented
in wide form for domain-expert review.

```text
semantic and review design
          ↓
     atomic claims
          ↓
   bounded review task
          ↓
    wide projection
          ↓
  domain-expert review
```

This is a division of cognitive work rather than a hierarchy of semantic
validity. The wide projection preserves the candidate semantic structure while
presenting related assertions in a form suited to the review task.

## Rendering the wide projection

Betwixt currently provides browser-based rendering for the wide projection.
The renderer adapts the candidate dataset to the review interface without
changing its semantic contract.

Evidence and descriptive information remain visible alongside the candidate
assertions. Depending on the candidate data, reviewable values can be presented
as resolved entities, editable unresolved entities, or controlled choices.
Definitions can link to semantic resources, while controlled ranges constrain
or guide the available values. Context remains visible but outside the review
boundary.

Reviewers can accept or modify candidate values, defer or reject assertions,
and finalise reviewed rows. The interface also records review metadata,
including the reviewer, project, review sequence, timestamps, and review
status.

A review can be saved as a draft and subsequently finalised. Draft and
finalised files from the same review cycle retain the same sequence number, so
the standalone HTML preserves both the review interface and its current review
state.

```{r delini-wide-render}
betwixt_render(
  delini,
  cols = c(
    col_1 = "Subject",
    col_2 = "instance of",
    col_3 = "heritage of",
    context_1 = "held by"
  ),
  subheadings = c(
    col_2 = "wdt:P31",
    col_3 = "controlled range"
  ),
  title = "Delini semantic review",
  description = "Review the proposed semantic assertions.",
  project_id = "delini",
  filename_stem = "delini-wide",
  sequence = 0L,
  path = tempdir()
)
```

The wide projection is the browser-rendered review interface in the current
Betwixt implementation. The other Delini projections demonstrate how the same
candidate knowledge can be rearranged or extended without making any one
tabular representation canonical.
