---
title: "Review Representations and Projections"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Review Representations and Projections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(betwixt)
```

# Review representations and projections

Betwixt separates the representation used to prepare a review task from the representations produced after human review. Candidate assertions are prepared in a wide tabular representation that can be created programmatically or imported from ordinary tabular data.

The review task itself can be designed in different ways. In a **wide** review, evidence supports assertions about a subject. In a **dual-wide** review, the relation between the evidence and the subject is also brought inside the review boundary.

After human review, the resulting review state can be projected into wide review planes or into long-form atomic semantic assertions.

```text
candidate dataset
       ↓
 review design
   ↙       ↘
wide     dual-wide
   ↘       ↙
  human review
       ↓
 reviewed state
   ↙       ↘
wide       long
projection projection
```

These distinctions serve different purposes. Wide and dual-wide concern **what the reviewer is asked to evaluate**. Wide and long projections concern **how the resulting review state is represented for subsequent use**.

## The Dēliņi example

The `delini` dataset is a small cultural heritage example associated with the Dēliņi farmstead at the Ethnographic Open-Air Museum of Latvia. It contains five observations concerning three heritage artefacts and two documentary records.

```{r delini}
data("delini")

delini
```

Each row combines several kinds of information required for a review task:

- **evidence** identifies or presents material supporting the observation;
- **description** provides human-readable information for the reviewer;
- **candidate columns** contain semantic assertions proposed for review;
- **context** provides useful information outside the current review boundary.

These roles are represented with ordinary columns rather than a specialised R object. This makes the candidate dataset suitable for exchange through tabular formats such as CSV and Excel.

## Candidate columns

A candidate column represents something that can be reviewed. In `delini`, the principal candidate columns are:

```text
subject
instance_of
heritage_of
```

A candidate column may be accompanied by two related columns:

```text
<name>
<name>_range
<name>_definition
```

For example:

```text
instance_of
instance_of_range
instance_of_definition
```

The candidate column contains the proposed value. Its `_range` column can provide controlled or suggested alternatives, while its `_definition` column can provide a semantic definition or reference associated with the candidate.

For the `instance_of` candidate in `delini`, these columns distinguish the value proposed for a particular observation from the range available to the reviewer and the semantic resource defining the relation.

In R, these columns can be added programmatically with `add_candidate_column()`, `add_candidate_range()` and `add_candidate_definition()`. The tabular convention itself is independent of R: the same structure can be prepared manually in CSV or Excel and imported as a candidate dataset.

## Evidence, description and context

Evidence is deliberately kept distinct from the candidate assertions being reviewed. The Dēliņi example can provide an evidence resource through `evidence_url`, directly displayable media through `evidence_media_url`, and a short identification through `evidence_text`.

The `label` and `description` columns help the reviewer understand the resource but are not automatically reviewable assertions.

Similarly, `context_1` provides information useful during review without bringing that information inside the current review boundary. In the Dēliņi example it identifies the institution holding the material.

This distinction allows a review task to present rich supporting information without treating everything visible to the reviewer as a candidate assertion.

## Wide review

The candidate dataset is deliberately wide. A single row keeps an observation, its evidence and several related candidate assertions together:

```text
Evidence | Subject | instance_of | heritage_of | Context
```

For domain experts this is often more natural than presenting every semantic assertion as a separate row. The reviewer can inspect the resource once and consider several proposed assertions about it together.

In an ordinary wide review, the evidence supports the review task but its relation to the subject is not itself reviewed.

```text
Evidence ──────► Subject | instance_of | heritage_of
  context           └──────── reviewable ──────────┘
```

### Rendering the wide review

`render_review()` transforms the candidate dataset into a standalone browser-based review artefact.

```{r delini-wide-render}
wide_review_file <- render_review(
  delini,
  cols = c(
    subject = "Subject",
    instance_of = "instance of",
    heritage_of = "heritage of",
    context_1 = "held by"
  ),
  subheadings = c(
    instance_of = "type",
    heritage_of = "cultural association"
  ),
  title = "Dēliņi semantic review",
  description = "Review the proposed semantic assertions.",
  project_id = "delini",
  filename_stem = "delini-wide",
  sequence = 0L,
  path = tempdir()
)
```

Rendering does not turn candidate values into accepted knowledge. It creates a human review environment in which proposed values can be corroborated, corrected, rejected or deferred.

The resulting HTML combines the candidate assertions with the evidence and descriptive information needed to evaluate them. It also records the review state and provenance required to read the review back into Betwixt.

## Dual-wide review

Sometimes the relationship between evidence and subject is itself uncertain or semantically important. A photograph may **depict** an artefact, while an archival record may **document** it.

A dual-wide review brings this relation inside the review boundary:

```text
Evidence ─relation─► Subject | instance_of | heritage_of
          └────────────── reviewable ──────────────┘
```

For the Dēliņi example, we can create a second candidate dataset in which the evidence relation is explicitly reviewable.

```{r delini-dual}
delini_dual <- create_candidate_dataset(
  evidence_url = delini$evidence_url,
  evidence_media_url = delini$evidence_media_url,
  evidence_text = delini$evidence_text,
  label = delini$label,
  description = delini$description,
  subject = delini$subject,
  subject_range = delini$subject_range,
  subject_definition = delini$subject_definition,
  evidence_relation = c(
    "depicts", "depicts", "depicts", "documents", "documents"
  ),
  evidence_relation_range = add_candidate_range(
    "depicts", "documents", "Other…"
  )
)

delini_dual <- delini_dual |>
  add_candidate_column(
    name = "instance_of",
    value = delini$instance_of,
    range = delini$instance_of_range,
    definition = delini$instance_of_definition
  ) |>
  add_candidate_column(
    name = "heritage_of",
    value = delini$heritage_of,
    range = delini$heritage_of_range,
    definition = delini$heritage_of_definition
  )
```

The two candidate datasets contain the same assertions about the heritage resources. They differ in the boundary of the review task.

In the dual-wide representation, `evidence_relation` is itself a candidate assertion. It remains distinct from review provenance, which records who performed the review, when the review took place, and what decisions were made.

### Rendering the dual-wide review

The same renderer is used for the dual-wide candidate dataset:

```{r delini-dual-render}
dual_review_file <- render_review(
  delini_dual,
  cols = c(
    evidence_relation = "Evidence relation",
    subject = "Subject",
    instance_of = "instance of",
    heritage_of = "heritage of",
    context_1 = "held by"
  ),
  title = "Dēliņi dual-wide semantic review",
  description = "Review the evidence relation and proposed assertions.",
  project_id = "delini",
  filename_stem = "delini-dual-wide",
  sequence = 0L,
  path = tempdir()
)
```

Wide and dual-wide are therefore not different semantic models. They are different **review designs**. The wide review treats evidence as supporting material. The dual-wide review additionally asks the reviewer to evaluate how that evidence relates to the subject.

## From review to reviewed state

After a review has been saved, `read_review()` reads the standalone review artefact back into Betwixt.

```r
review <- read_review("delini-wide-finalised.html")
```

The resulting review object preserves the distinction between what was originally proposed and what resulted from human review.

The reviewed state can subsequently be projected in two useful forms:

```text
                    ┌─ wide → candidate / reviewed / status
reviewed state ─────┤
                    └─ long → atomic semantic assertions
```

These are projections of the same reviewed state rather than alternative review layouts.

## Wide review projection

`project_review_wide()` exposes three related planes of the review:

```r
wide <- project_review_wide(review)
```

The **candidate** plane records the values originally proposed for review. The **reviewed** plane records the values resulting from human review. The **status** plane records the review outcome associated with each reviewable value.

Conceptually:

```text
candidate   proposed value
reviewed    value after human review
status      review outcome
```

Keeping these planes distinct makes changes introduced through review explicit rather than overwriting the original candidate material.

## Long assertion projection

`project_review_long()` projects the reviewed state into atomic semantic assertions:

```r
long <- project_review_long(review)
```

Instead of keeping several candidate predicates on one wide row, the long projection represents individual assertions explicitly:

```text
subject | predicate | value | status
```

A wide review row can therefore produce several atomic assertions. The long representation makes the subject–predicate–value structure explicit and is suited to subsequent semantic processing and serialisation.

The long projection does not introduce new knowledge. It projects the reviewed state into a representation in which the individual assertions and their review status are explicit.

## Review design and semantic projection

Betwixt therefore distinguishes two operations that can otherwise look superficially similar.

**Review design** determines what is presented to the reviewer:

```text
wide       evidence supports the reviewed assertions
dual-wide  evidence relation is itself reviewable
```

**Projection** determines how the resulting review state is represented:

```text
wide       candidate, reviewed and status planes
long       atomic subject–predicate–value assertions
```

This separation allows the review interface to remain suited to human judgement while the resulting reviewed knowledge can be projected into forms suited to subsequent computational use.

```text
candidate assertions
        ↓
    review design
        ↓
    human review
        ↓
   reviewed state
        ↓
semantic projection
        ↓
subsequent processing
and serialisation
```
